---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rainonsnow

<!-- badges: start -->
<!-- badges: end -->

`rainonsnow` is a research repository for exploring rain-on-snow hydrology in the Alps, focussing on extreme runoff as a consequence.

The repo consists of:

- A code pipeline for bringing in raster data from Google Earth Engine and analyzing it.
- A set of Shiny apps for exploring the data and models.
- An R package for the distributional learning model and helper functions.

## Current Workflow

Scripts are numbered in pipeline order (`scripts/1-*` … `scripts/7-*`). Intermediate files land under `derived/` (and yearly NetCDF under `derived/eo/`).

1. **`scripts/1-download_data-eo.py`** — Download hourly ERA5-Land from Google Earth Engine to `derived/eo/era5_land_hourly_alps_<year>.nc` (bands and years from `inputs/data_specifications.yaml`).
2. **`scripts/2-tablify_spatial_eo.r`** — Raster NetCDF → tabular hourly series; writes `derived/era5_land_hourly_alps_all.rds`.
3. **`scripts/3-pot_spatial_eo.r`** — Peaks over threshold per cell (`derived/era5_land_hourly_alps_peaks.rds`, thresholds, metadata).
4. **`scripts/4-distributional_learning.r`** — Quantile regression forest on POT peak hours (models and predictions used by later apps).
5. **`scripts/5-runoff_marginals.r`** — Marginal return-level tables for runoff (`derived/era5_land_hourly_alps_dl_*`).
6. **`scripts/6-drivers_joint_distribution.r`** — Joint rainfall–snowmelt model per cell.
7. **`scripts/7-likeliest_rain_snow.r`** — Conditional rain–snow structure given extreme runoff (optional follow-on).

## Dependency management

Python dependencies are managed with **[uv](https://github.com/astral-sh/uv)** (`pyproject.toml`; run `uv sync` so the lockfile matches the resolved environment). R package dependencies are declared in `DESCRIPTION`; install them from the repo root with something like `devtools::install_deps()` (or your preferred workflow). While editing package code, `devtools::load_all()` attaches the package from source.

## Python Setup

This repository uses `uv` for the Python environment.

```bash
uv sync
```

The Python dependencies are declared in `pyproject.toml`.

## Data specifications (`inputs/data_specifications.yaml`)

Pipeline settings for ERA5-Land export and tabular cleanup live in **`inputs/data_specifications.yaml`**:

- **`earth_engine.project_id`** — Google Cloud project passed to `ee.Initialize(project=...)`.
- **`download`** — Earth Engine **`collection_id`**, **`first_year`** / **`last_year`**, and **`variables`** (hourly band names exported from the collection).
- **`tablify.epsilon_mm`** — Near-zero cutoff for rain and melt (mm/h equivalent) in script 2 before other steps.

Authentication is separate from this file: run once per machine/user:

```bash
earthengine authenticate
```

Run the downloader with:

```bash
uv run python scripts/1-download_data-eo.py
```

If `data_specifications.yaml` is missing, scripts 1 and 2 fall back to the same defaults as the template file.

## Input Controls

YAML files under **`inputs/`** configure scripts and apps (beyond `data_specifications.yaml` above):

- **`distributional_learning.yaml`** — Response and predictors for **`dl_rqforest`** in script 4.
- **`rain_snow_joint_model.yaml`** — Joint marginal + copula options for script 6 (key `fit_joint_rain_snow_cells`), and **`likeliest_rain_snow`** settings for script 7 (conditional rain–snow surface given runoff).
- **`pot_metadata.yaml`** — POT quantile / min-gap for script 3 and `apps/3-pot-explorer`.

## R Setup

In addition to the analysis stream of this repository, this repository is also structured like an R package, with source files in `R/`, documentation in `man/`, and package metadata in `DESCRIPTION` and `NAMESPACE`. This is useful so that custom functions can be more easily accessed by the analysis scripts, and more easily used by anyone developing this analysis.

To make these functions available to the analysis scripts, run:

```r
devtools::load_all()
```

If you make a change to the R package, run `devtools::document()` to update the package documentation.

## Run The Analysis

Run the scripts in order (each writes inputs for the next) from the repo root.

```bash
uv python scripts/1-download_data-eo.py
Rscript scripts/2-tablify_spatial_eo.r
Rscript scripts/3-pot_spatial_eo.r
Rscript scripts/4-distributional_learning.r
Rscript scripts/5-runoff_marginals.r
Rscript scripts/6-drivers_joint_distribution.r
```

## Tail modelling

The marginal runoff distribution for a cell is the equal-weight mixture of its
peak-hour predictive distributions, each grafted to a generalized Pareto tail.
That construction has a trap. If component $i$ has shape $\xi_i$, its survival
is regularly varying with index $1/\xi_i$, so the mixture's index is
$\min_i (1/\xi_i)$ — the cell's tail is set by $\max_i \xi_i$ alone. Each
$\xi_i$ is fitted to the upper half of one predictive distribution, whose
effective sample size is small, so that maximum is the maximum of a few hundred
noisy estimates.

The fix is to share the shape across a cell and let only the scale vary by hour,
which is also the standard non-stationary POT model: covariates move the scale,
not the index. `fit_gpd_shared_shape()` profiles the shared shape, optimising
each component's scale by a bounded one-dimensional search. `mixture_tail()`
then evaluates the mixture in closed form.

`scripts/experiments/tail-index-pooling.R` measures all of this on simulated
cells where the truth is known (results in
`scripts/experiments/results/`). With 150 peak hours and 15 effective tail
points each, at a true $\xi$ of 0.15:

| shape estimator | effective $\xi$ | T=10y | T=100y | T=1000y |
|---|---|---|---|---|
| one per hour (previous behaviour) | 0.93 | 1.00x | 1.02x | **1.64x** |
| shared across the cell | 0.07 | 1.00x | 0.99x | 0.96x |
| shared, shape bias-corrected | 0.17 | 1.07x | 1.16x | 1.24x |
| true shape supplied | 0.15 | 1.01x | 1.07x | 1.10x |

Columns are the median ratio to the correct return level. Two things are worth
noting. Fitting a shape per hour is not merely noisy, it diverges as the return
period grows, which is exactly the regime the project cares about. And the row
with the true shape supplied does not sit at 1.00x either: each hour's *scale*
is fitted from a handful of points and the return level is convex in those
scales, so scale noise inflates it by 7–10% on its own. That second error is
still open — see `?fit_gpd_shared_shape` for why the shape bias correction is
off by default as a result.

Shapes can also be smoothed between neighbouring cells
(`smooth_tail_shape()`), by empirical-Bayes shrinkage toward the neighbourhood
mean. Part 2 of the same experiment checks both the benefit and the risk:

| true shape field | per-cell | neighbour-smoothed | single global shape |
|---|---|---|---|
| flat | 0.053 | 0.027 | **0.013** |
| gentle trend | 0.058 | **0.030** | 0.033 |
| steep trend | 0.055 | **0.041** | 0.086 |

RMSE in $\xi$. Neighbour smoothing helps in every regime and never hurts,
whereas collapsing to one global shape is best when the field is genuinely flat
and clearly harmful when it is not.

### Evaluating the mixture

Above `max(graft_of)` every component is in its GP part, so the mixture survival
is elementary in the stored `(graft_of, graft_tail_prob, gp_scale, gp_shape)`
columns — no distribution objects are built at all. `mixture_tail_compress()`
further bins components by scale while preserving their combined asymptotic
contribution (5e-5 relative error at 64 bins). A 500-point return curve takes
about 50 ms, which is what makes `scripts/7-*-animated.r` practical.

Below `max(graft_of)` the closed form does not apply and evaluation returns `NA`
rather than extrapolating the GP into a region where the mixture still has
empirical mass; script 5 falls back to the body mixture there and records which
region each return level came from.

## Copula transport of the marginal

The equal-weight mixture of the fitted conditionals is not the only way to reach
the marginal of runoff, and it turns out not to be an unbiased one.

Write $U = F_X(X)$, $V = F_Y(Y)$ and $W = 1/(1-V)$, which is standard Pareto
marginally. The **copula conditional EVI**, $\mathrm{CEVI}(u)$, is the extreme
value index of $W$ given $U = u$: a property of the copula alone. For heavy or
light tailed $Y$,

$$\mathrm{EVI}(Y \mid X = x) \;=\; \mathrm{CEVI}(u)\cdot \mathrm{EVI}(Y), \qquad u = F_X(x),$$

with $\mathrm{CEVI} \in [0,1]$. Every conditional is lighter-tailed than the
marginal, and the copula says by exactly how much. For the Gaussian copula
$\mathrm{CEVI} = 1-\rho^2$, constant in $u$ (verified numerically in part 1 of
the experiment): the stronger the dependence, the more of the marginal's tail
heaviness is explained by the predictor rather than by conditional noise.

### The mixture marginal is biased low, and more data does not fix it

Every component of the mixture shares the conditional EVI, so the mixture is
regularly varying with that index -- which is $\mathrm{CEVI}\cdot\mathrm{EVI}(Y)$,
strictly lighter than the truth whenever $\mathrm{CEVI} < 1$. The true marginal
only acquires its full heaviness from covariate values beyond any finite sample.
So the deficit is structural, not an estimation error, and it does not shrink
with $n$ (`scripts/experiments/copula-transport-sweep.R`; median ratio to the
true return level at $p = 10^{-4}$):

| $\rho$ | CEVI | n = 300 | n = 1000 | n = 3000 |
|---|---|---|---|---|
| 0.5 | 0.75 | 0.84x | 0.88x | 0.89x |
| 0.7 | 0.51 | 0.67x | 0.75x | 0.78x |
| 0.9 | 0.19 | 0.55x | 0.59x | 0.61x |

At dependence strong enough to be worth modelling, the mixture understates the
10,000-event return level by 20--40%, and tripling the sample barely moves it.

### Transport fixes it, where the conditional index can be estimated

Pushing a conditional back through the copula,
$F_Y(y) = h^{-1}\!\left(F_{Y|X=x}(y) \mid u\right)$ with $h(v|u) = \partial C(u,v)/\partial u$,
recovers the marginal exactly, and splits the estimation into a conditional tail
index (from all $n$ points, at a shallower extrapolation) and a copula parameter
(from all $n$ ranks, not just the tail). With a local conditional model
(part 4 of the experiment), at $\rho = 0.7$ and $n = 3000$, $p = 10^{-4}$:

| estimator | median | RMSE(log) |
|---|---|---|
| mixture over observed covariates | 0.77x | 0.28 |
| copula transport | **0.97x** | **0.26** |
| transport, anchored on the marginal body | 0.94x | 0.25 |

Better on bias *and* on total error.

### Two things that decide whether it works

**The conditional model does most of the work.** Fitting the conditionals by
binning the covariate makes the answer depend on the bin width: 5 bins at
n = 3000 inflates the conditional EVI to 0.135 against a true 0.128 (each bin is
a mixture over its own covariate range) and the transport comes out 20% high,
while 10--20 bins recovers 0.107--0.110 and lands within 10%. Overlapping
nearest-neighbour neighbourhoods -- much closer to what the quantile regression
forest actually does -- are what the table above uses.

**The decomposition is ill-conditioned when CEVI is small.** Since
$\mathrm{EVI}(Y) = \mathrm{EVI}(Y|X)/\mathrm{CEVI}$, any error in the conditional
index is amplified by $1/\mathrm{CEVI}$. At $\rho = 0.9$ that factor is 5.3, the
conditional tail is nearly exponential and its index estimates negative, and
every method fails. So the transport is most attractive in the middle: strong
enough dependence that the mixture is visibly biased, not so strong that the
conditional tail carries no information.

A null check is included: when the covariate effect is bounded so that
$\mathrm{CEVI} = 1$ and the copula explains none of the tail heaviness, the
anchored transport neither gains nor does harm (1.04x at $p = 10^{-4}$).

## Shiny apps

Interactive apps live under `apps/*`, with folder names numbered to match the scripts they support. Run them from the **repository root** so paths such as `derived/` and `devtools::load_all()` resolve:

```r
shiny::runApp("apps/<app-folder>")
```

The apps share a look and feel through `apps/ros_theme.R` (Bootstrap 5 via
**bslib**, plus a common ggplot theme and palette); if bslib is not installed
they fall back to default styling rather than failing to start.

Typical dependencies include [shiny](https://shiny.posit.co/), **ggplot2**, **tidyverse**, **sf**, and **rnaturalearth**; individual apps may need **yaml**, **probaverse**, **distionary**, **famish**, **rvinecopulib**, and others used in the analysis scripts.

### POT explorer (`apps/3-pot-explorer`)

Inspect **peaks over threshold (POT)** extraction: hourly runoff for one grid cell and year with the POT threshold and marked peaks; **event timing** scatter of all POT peaks for that cell (runoff vs day of year, all years pooled). Pick the cell from the map or sidebar; optional re-run of script 3 from the sidebar.

**Data:** `derived/era5_land_hourly_alps_all.rds` (`scripts/2-tablify_spatial_eo.r`) and `derived/era5_land_hourly_alps_peaks.rds` (`scripts/3-pot_spatial_eo.r`).

```r
shiny::runApp("apps/3-pot-explorer")
```

### Distributional learning — fit (`apps/4a-distributional-learning-fit`)

Tune **`dl_rqforest`**, preview on one cell, then run script 4 for all cells (writes **`inputs/distributional_learning.yaml`** and derived outputs).

```r
shiny::runApp("apps/4a-distributional-learning-fit")
```

### Distributional learning — diagnostics (`apps/4b-distributional-learning-diagnostics`)

P–P, skill, map, and rain–snow CDF exploration. Models load at startup; load calibration diagnostics on demand (reads `derived/era5_land_hourly_alps_dl_diagnostics.rds` from script 4 when present).

```r
shiny::runApp("apps/4b-distributional-learning-diagnostics")
```

### Return-level explorer (`apps/5-return-level-explorer`)

Map of marginal runoff return levels by cell, frequency–magnitude curves (forest mixture vs GP tail), and rain–snow likelihood surfaces at a chosen return period.

**Data:** peaks from script 3; script 4 models; precomputed marginal return levels from script 5 (`derived/era5_land_hourly_alps_dl_marginal_return_levels.rds` or the bundle described in the app header), or `derived/era5_land_hourly_alps_dl_predictions.rds` as a slower fallback.

```r
shiny::runApp("apps/5-return-level-explorer")
```

### Tail shape explorer (`apps/5b-tail-shape-explorer`)

Map of the shared tail index per cell, how far neighbour smoothing moved each
one, and how much the design return level actually depends on the shape.

**Data:** `derived/era5_land_hourly_alps_dl_tail_summary.rds` and
`derived/era5_land_hourly_alps_dl_mixture_tails.rds` (script 5); optionally
`derived/era5_land_hourly_alps_dl_tail_shapes.rds` (script 4).

``` r
shiny::runApp("apps/5b-tail-shape-explorer")
```

### Joint rainfall–snowmelt explorer (`apps/6-joint-rain-snow-explorer`)

Marginal fits and copula per cell: Gaussian-score diagnostics, joint density contours, marginal histograms, and frequency–magnitude curves for rainfall and snowmelt. Optional re-run of joint fitting from the sidebar (writes `inputs/rain_snow_joint_model.yaml` and runs `scripts/6-drivers_joint_distribution.r`).

**Data:** `derived/era5_land_hourly_alps_all.rds` and joint output from script 6 (`derived/era5_land_hourly_alps_joint_rain_snow.rds`); fitting options and script 7 settings share `inputs/rain_snow_joint_model.yaml`.

```r
shiny::runApp("apps/6-joint-rain-snow-explorer")
```

### Runoff marginals explorer (`apps/5-runoff-marginals-explorer`)

Side-by-side frequency–magnitude curves: distributional-learning marginals (Random Forest mixture vs GP conversion) and **naive** POT-only marginals (`distionary::dst_empirical` vs `famish::fit_dst_gp` on peak runoff). Map cell selection; optional matched *y*-axis limits and log return level on both panels.

**Data:** `derived/era5_land_hourly_alps_peaks.rds` and `derived/era5_land_hourly_alps_dl_return_levels.rds` from `scripts/5-runoff_marginals.r`.

```r
shiny::runApp("apps/5-runoff-marginals-explorer")
```

### Rain and snowmelt given extreme runoff (`apps/7-rain-snow-given-runoff`)

Interactive version of `scripts/7-likeliest_rain_snow-one_cell_animated.r`: the
return period is a slider with a play button, so T can be scrubbed as well as
animated. Each grid point's predictive tail is fitted once per cell and every
frame is one vectorised density evaluation, which is what makes it interactive
at all.

**Data:** peaks (script 3), fitted models (script 4),
`derived/era5_land_hourly_alps_dl_mixture_tails.rds` (script 5) and the joint
model (script 6).

``` r
shiny::runApp("apps/7-rain-snow-given-runoff")
```

## R Package Demonstration

```{r}
library(rainonsnow)
library(dplyr)
library(probaverse)
```

The R package is titled `rainonsnow` and provides a small distributional learning interface for modelling the distribution of a target variable given some predictors.

For a simple workflow, fit a model with `dl_rqforest()` and then call `predict()`, using the `mtcars` dataset from the stats package.

```{r}
df <- as_tibble(mtcars)
model <- dl_rqforest(
  data = df,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)

df <- mutate(df, distribution = predict(model), .before = everything())
df
```

The predictions are distributions. Take a look at the first distribution using the probaverse, for example, and plot its cdf.

```{r}
plot(df$distribution[[1]], n = 1000)
```

The package also includes a null model to handle failures gracefully. For example, if you ask for `na_action = "null"` and the training data contain missing values, or if the training fails, `dl_rqforest()` returns a `dl_null` object instead of failing:

```{r}
df2 <- as_tibble(mtcars)
df2$wt[2] <- NA_real_

dl_rqforest(
  data = df2,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)
```

Predicting on these objects always returns a null distribution with the number of rows of the data:

```{r}
predict(dl_null(), newdata = tibble(x = 1:2))
```

## Status

This repository is in an active research/prototyping state, so scripts, paths, and interfaces may change as the workflow evolves.

## License

This repository is licensed under the MIT License - see the `LICENSE` file for details.