---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rainonsnow

<!-- badges: start -->
<!-- badges: end -->

`rainonsnow` is a research repository for exploring rain-on-snow hydrology in the Alps, focussing on extreme runoff as a consequence.

The repo consists of:

- A code pipeline for bringing in raster data from Google Earth Engine and analyzing it.
- A set of Shiny apps for exploring the data and models.
- An R package for the distributional learning model and helper functions.

## Current Workflow

The work splits into **shared data** (built once) and **analyses** (many, each in its own folder).

Shared data, in `derived/`:

1. **`scripts/1-download_data-eo.py`** — Download hourly ERA5-Land from Google Earth Engine to `derived/eo/era5_land_hourly_alps_<year>.nc` (bands and years from `inputs/data_specifications.yaml`).
2. **`scripts/2-tablify_spatial_eo.r`** — Raster NetCDF → tabular hourly series; writes `derived/era5_land_hourly_alps_all.rds`.
3. **`scripts/3-pot_spatial_eo.r`** — Peaks over threshold per cell (`derived/era5_land_hourly_alps_peaks.rds`, thresholds).

Analyses, in `analyses/<name>/`: each one chooses predictors, a distributional learning model, and queries in its `analysis.yaml`, and is run with

```bash
Rscript scripts/run-analysis.R <name>
```

See **[`analyses/README.md`](analyses/README.md)** for the index of analyses, the stages, and how to add one. Browse all of them in one app:

```r
shiny::runApp("apps/explorer")
```

## Dependency management

Python dependencies are managed with **[uv](https://github.com/astral-sh/uv)** (`pyproject.toml`; run `uv sync` so the lockfile matches the resolved environment). R package dependencies are declared in `DESCRIPTION`; install them from the repo root with something like `devtools::install_deps()` (or your preferred workflow). While editing package code, `devtools::load_all()` attaches the package from source.

## Python Setup

This repository uses `uv` for the Python environment.

```bash
uv sync
```

The Python dependencies are declared in `pyproject.toml`.

## Data specifications (`inputs/data_specifications.yaml`)

Pipeline settings for ERA5-Land export and tabular cleanup live in **`inputs/data_specifications.yaml`**:

- **`earth_engine.project_id`** — Google Cloud project passed to `ee.Initialize(project=...)`.
- **`download`** — Earth Engine **`collection_id`**, **`first_year`** / **`last_year`**, and **`variables`** (hourly band names exported from the collection).
- **`tablify.epsilon_mm`** — Near-zero cutoff for rain and melt (mm/h equivalent) in script 2 before other steps.

Authentication is separate from this file: run once per machine/user:

```bash
earthengine authenticate
```

Run the downloader with:

```bash
uv run python scripts/1-download_data-eo.py
```

If `data_specifications.yaml` is missing, scripts 1 and 2 fall back to the same defaults as the template file.

## Input Controls

- **`inputs/data_specifications.yaml`** — download and tablify settings (scripts 1-2).
- **`inputs/pot_metadata.yaml`** — POT quantile / min-gap for script 3 and `apps/3-pot-explorer`.
- **`analyses/<name>/analysis.yaml`** — everything model-specific (predictors, model, tail, queries).

## R Setup

In addition to the analysis stream of this repository, this repository is also structured like an R package, with source files in `R/`, documentation in `man/`, and package metadata in `DESCRIPTION` and `NAMESPACE`. This is useful so that custom functions can be more easily accessed by the analysis scripts, and more easily used by anyone developing this analysis.

To make these functions available to the analysis scripts, run:

```r
devtools::load_all()
```

If you make a change to the R package, run `devtools::document()` to update the package documentation.

## Run The Analysis

From the repo root:

```bash
uv run python scripts/1-download_data-eo.py
Rscript scripts/2-tablify_spatial_eo.r
Rscript scripts/3-pot_spatial_eo.r
Rscript scripts/run-analysis.R rain-llqr
Rscript scripts/run-analysis.R rain-snowmelt-rqforest
```

## Shiny apps

Run from the **repository root**.

- **`apps/explorer`** — every analysis in one place: pick the analysis, cell and T-year event; tabs for the model (including the local-fit picture for `llqr`), trigger rain, likeliest drivers, a cross-analysis comparison, and diagnostics.
- **`apps/3-pot-explorer`** — inspect the shared POT extraction: hourly runoff with threshold and peaks for one cell and year, and event timing.

## R Package Demonstration

```{r}
library(rainonsnow)
library(dplyr)
library(probaverse)
```

The R package is titled `rainonsnow` and provides a small distributional learning interface for modelling the distribution of a target variable given some predictors.

For a simple workflow, fit a model with `dl_rqforest()` and then call `predict()`, using the `mtcars` dataset from the stats package.

```{r}
df <- as_tibble(mtcars)
model <- dl_rqforest(
  data = df,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)

df <- mutate(df, distribution = predict(model), .before = everything())
df
```

The predictions are distributions. Take a look at the first distribution using the probaverse, for example, and plot its cdf.

```{r}
plot(df$distribution[[1]], n = 1000)
```

The package also includes a null model to handle failures gracefully. For example, if you ask for `na_action = "null"` and the training data contain missing values, or if the training fails, `dl_rqforest()` returns a `dl_null` object instead of failing:

```{r}
df2 <- as_tibble(mtcars)
df2$wt[2] <- NA_real_

dl_rqforest(
  data = df2,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)
```

Predicting on these objects always returns a null distribution with the number of rows of the data:

```{r}
predict(dl_null(), newdata = tibble(x = 1:2))
```

## Status

This repository is in an active research/prototyping state, so scripts, paths, and interfaces may change as the workflow evolves.

## License

This repository is licensed under the MIT License - see the `LICENSE` file for details.