---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rainonsnow

<!-- badges: start -->
<!-- badges: end -->

`rainonsnow` is a research repository for exploring rain-on-snow hydrology in the Alps, focussing on extreme runoff as a consequence.

The repo currently has two main pieces:

- A Python workflow for downloading hourly ERA5-Land variables from Google Earth Engine.
- An R package plus analysis script for fitting and inspecting distributional learning models on the exported data.

## Current Workflow

1. Download gridded hydroclimate data with `scripts/download_data.py`.
2. Save the resulting NetCDF file locally.
3. Analyze the data and fit models from R with `scripts/analyze.r`.

The Python downloader writes `data/era5_land_hourly_alps.nc`, which matches the path used by `scripts/analyze.r`.

## Python Setup

This repository uses `uv` for the Python environment.

```bash
uv sync
```

The Python dependencies are declared in `pyproject.toml`.

## Download Data

The downloader script initializes Earth Engine, queries the `ECMWF/ERA5_LAND/HOURLY` image collection for an Alps bounding box, selects a small set of snow, precipitation, temperature, and runoff variables, and exports the result to NetCDF.

Before running it, make sure you have authenticated with Google Earth Engine and that the project in `scripts/download_data.py` is valid for your account:

```bash
earthengine authenticate
```

Run the script with:

```bash
uv run python scripts/download_data.py
```

## R Setup

The R side of the repository is structured like a package, with source files in `R/`, documentation in `man/`, and package metadata in `DESCRIPTION` and `NAMESPACE`.

Install the required R packages as needed, then load the package locally during development:

```r
devtools::load_all()
```

## Run The Analysis

Open `scripts/analyze.r` in R and run it interactively, or source it from an R session.

The script currently:

- Loads the NetCDF file with `terra::rast()`
- Converts the raster stack into a tabular format
- Fits a distributional learning model per cell
- Generates predictions from the fitted models

## POT explorer (Shiny)

There is a small interactive app under `apps/pot-explorer/` for inspecting **peaks over threshold (POT)** extraction: hourly runoff for one grid cell and year, the POT threshold line, and the peak events that enter the POT sample. You can pick the cell from a map or from the sidebar.

**Prerequisites:** install [shiny](https://shiny.posit.co/), **ggplot2**, **dplyr**, **sf**, and **rnaturalearth** if you do not already have them (the same kinds of packages used in `scripts/analyze.qmd` for maps).

**Data:** the app reads tabulated hourly series and POT peaks from the project `data/` folder:

- `data/era5_land_hourly_alps_all.rds` — produced by `scripts/2-tablify_spatial_eo.r`
- `data/era5_land_hourly_alps_peaks.rds` — produced by `scripts/3-pot_spatial_eo.r`

Run those scripts first so both files exist (the first load can take a while if the hourly RDS is large).

**Launch** from the repository root in R:

```r
shiny::runApp("apps/pot-explorer")
```

The working directory must be the app folder’s parent (`rainonsnow/`), which is the usual behaviour when you call `runApp()` with that path from an R session whose working directory is the repo root.

## R Package Contents

```{r}
library(rainonsnow)
library(dplyr)
```

The R package part of the repository provides a small distributional learning interface for modelling the distribution of a target variable given some predictors.

For a simple workflow, fit a model with `dl_rqforest()` and then call `predict()`, using the `mtcars` dataset from the stats package.

```{r}
df <- as_tibble(mtcars)
model <- dl_rqforest(
  data = df,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)

df <- mutate(df, distribution = predict(model), .before = everything())
df
```

The predictions are distributions. Take a look at the first distribution using the probaverse, for example, and plot its cdf.

```{r}
plot(df$distribution[[1]], n = 1000)
```

The package also includes a null model to handle failures gracefully. For example, if you ask for `na_action = "null"` and the training data contain missing values, or if the training fails, `dl_rqforest()` returns a `dl_null` object instead of failing:

```{r}
df2 <- as_tibble(mtcars)
df2$wt[2] <- NA_real_

dl_rqforest(
  data = df2,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)
```

Predicting on these objects always returns a null distribution with the number of rows of the data:

```{r}
predict(dl_null(), newdata = tibble(x = 1:2))
```

## Repository Layout

```text
.
├── R/
├── man/
├── apps/
│   └── pot-explorer/
├── scripts/
│   ├── analyze.r
│   └── download_data.py
├── data/
├── main.py
├── DESCRIPTION
├── NAMESPACE
└── pyproject.toml
```

## Status

This repository is in an active research/prototyping state, so scripts, paths, and interfaces may change as the workflow evolves.