---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rainonsnow

<!-- badges: start -->
<!-- badges: end -->

`rainonsnow` is a research repository for exploring rain-on-snow hydrology in the Alps, focussing on extreme runoff as a consequence.

The repo consists of:

- A code pipeline for bringing in raster data from Google Earth Engine and analyzing it.
- A set of Shiny apps for exploring the data and models.
- An R package for the distributional learning model and helper functions.

## Current Workflow

Scripts are numbered in pipeline order (`scripts/1-*` … `scripts/7-*`). Intermediate files land under `data/` (and yearly NetCDF under `data/eo/`).

1. **`scripts/1-download_data-eo.py`** — Download hourly ERA5-Land from Google Earth Engine to `data/eo/era5_land_hourly_alps_<year>.nc` (bands and years from `inputs/data_specifications.yaml`).
2. **`scripts/2-tablify_spatial_eo.r`** — Raster NetCDF → tabular hourly series; writes `data/era5_land_hourly_alps_all.rds`.
3. **`scripts/3-pot_spatial_eo.r`** — Peaks over threshold per cell (`data/era5_land_hourly_alps_peaks.rds`, thresholds, metadata).
4. **`scripts/4-distributional_learning.r`** — Quantile regression forest on POT peak hours (models and predictions used by later apps).
5. **`scripts/5-runoff_marginals.r`** — Marginal return-level tables for runoff (`data/era5_land_hourly_alps_dl_*`).
6. **`scripts/6-drivers_joint_distribution.r`** — Joint rainfall–snowmelt model per cell.
7. **`scripts/7-likeliest_rain_snow.r`** — Conditional rain–snow structure given extreme runoff (optional follow-on).

## Dependency management

Python dependencies are managed with **[uv](https://github.com/astral-sh/uv)** (`pyproject.toml`; run `uv sync` so the lockfile matches the resolved environment). R package dependencies are declared in `DESCRIPTION`; install them from the repo root with something like `devtools::install_deps()` (or your preferred workflow). While editing package code, `devtools::load_all()` attaches the package from source.

## Python Setup

This repository uses `uv` for the Python environment.

```bash
uv sync
```

The Python dependencies are declared in `pyproject.toml`.

## Data specifications (`inputs/data_specifications.yaml`)

Pipeline settings for ERA5-Land export and tabular cleanup live in **`inputs/data_specifications.yaml`**:

- **`earth_engine.project_id`** — Google Cloud project passed to `ee.Initialize(project=...)`.
- **`download`** — Earth Engine **`collection_id`**, **`first_year`** / **`last_year`**, and **`variables`** (hourly band names exported from the collection).
- **`tablify.epsilon_mm`** — Near-zero cutoff for rain and melt (mm/h equivalent) in script 2 before other steps.

Authentication is separate from this file: run once per machine/user:

```bash
earthengine authenticate
```

Run the downloader with:

```bash
uv run python scripts/1-download_data-eo.py
```

If `data_specifications.yaml` is missing, scripts 1 and 2 fall back to the same defaults as the template file.

## Input Controls

YAML files under **`inputs/`** configure scripts and apps (beyond `data_specifications.yaml` above):

- **`distributional_learning.yaml`** — Response and predictors for **`dl_rqforest`** in script 4.
- **`rain_snow_joint_model.yaml`** — Joint marginal + copula options for script 6 (key `fit_joint_rain_snow_cells`), and **`likeliest_rain_snow`** settings for script 7 (conditional rain–snow surface given runoff).
- **`pot_metadata.yaml`** — POT quantile / min-gap for script 3 and `apps/pot-explorer`.

## R Setup

In addition to the analysis stream of this repository, this repository is also structured like an R package, with source files in `R/`, documentation in `man/`, and package metadata in `DESCRIPTION` and `NAMESPACE`. This is useful so that custom functions can be more easily accessed by the analysis scripts, and more easily used by anyone developing this analysis.

To make these functions available to the analysis scripts, run:

```r
devtools::load_all()
```

If you make a change to the R package, run `devtools::document()` to update the package documentation.

## Run The Analysis

Run the scripts in order (each writes inputs for the next) from the repo root.

```bash
uv python scripts/1-download_data-eo.py
Rscript scripts/2-tablify_spatial_eo.r
Rscript scripts/3-pot_spatial_eo.r
Rscript scripts/4-distributional_learning.r
Rscript scripts/5-runoff_marginals.r
Rscript scripts/6-drivers_joint_distribution.r
```

## Shiny apps

Interactive apps live under `apps/*`. Run them from the **repository root** so paths such as `data/` and `devtools::load_all()` resolve:

```r
shiny::runApp("apps/<app-folder>")
```

Typical dependencies include [shiny](https://shiny.posit.co/), **ggplot2**, **tidyverse**, **sf**, and **rnaturalearth**; individual apps may need **yaml**, **probaverse**, **distionary**, **famish**, **rvinecopulib**, and others used in the analysis scripts.

### POT explorer (`apps/pot-explorer`)

Inspect **peaks over threshold (POT)** extraction: hourly runoff for one grid cell and year with the POT threshold and marked peaks; **event timing** scatter of all POT peaks for that cell (runoff vs day of year, all years pooled). Pick the cell from the map or sidebar; optional re-run of script 3 from the sidebar.

**Data:** `data/era5_land_hourly_alps_all.rds` (`scripts/2-tablify_spatial_eo.r`) and `data/era5_land_hourly_alps_peaks.rds` (`scripts/3-pot_spatial_eo.r`).

```r
shiny::runApp("apps/pot-explorer")
```

### DL diagnostics explorer (`apps/dl-diagnostics-explorer`)

Calibration (P–P), skill versus marginal fit, and rain–snow scatter with conditional runoff CDFs at clicked coordinates.

**Data:** POT peaks, distributional-learning predictions, and fitted models from script 4 — e.g. `data/era5_land_hourly_alps_peaks.rds`, `data/era5_land_hourly_alps_dl_predictions.rds`, `data/era5_land_hourly_alps_dl_rqforest_models.rds`.

```r
shiny::runApp("apps/dl-diagnostics-explorer")
```

### Return-level explorer (`apps/return-level-explorer`)

Map of marginal runoff return levels by cell, frequency–magnitude curves (forest mixture vs GP tail), and rain–snow likelihood surfaces at a chosen return period.

**Data:** peaks from script 3; script 4 models; precomputed marginal return levels from script 5 (`data/era5_land_hourly_alps_dl_marginal_return_levels.rds` or the bundle described in the app header), or `data/era5_land_hourly_alps_dl_predictions.rds` as a slower fallback.

```r
shiny::runApp("apps/return-level-explorer")
```

### Joint rainfall–snowmelt explorer (`apps/joint-rain-snow-explorer`)

Marginal fits and copula per cell: Gaussian-score diagnostics, joint density contours, marginal histograms, and frequency–magnitude curves for rainfall and snowmelt. Optional re-run of joint fitting from the sidebar (writes `inputs/rain_snow_joint_model.yaml` and runs `scripts/6-drivers_joint_distribution.r`).

**Data:** `data/era5_land_hourly_alps_all.rds` and joint output from script 6 (`data/era5_land_hourly_alps_joint_rain_snow.rds`); fitting options and script 7 settings share `inputs/rain_snow_joint_model.yaml`.

```r
shiny::runApp("apps/joint-rain-snow-explorer")
```

### Runoff marginals explorer (`apps/runoff-marginals-explorer`)

Side-by-side frequency–magnitude curves: distributional-learning marginals (Random Forest mixture vs GP conversion) and **naive** POT-only marginals (`distionary::dst_empirical` vs `famish::fit_dst_gp` on peak runoff). Map cell selection; optional matched *y*-axis limits and log return level on both panels.

**Data:** `data/era5_land_hourly_alps_peaks.rds` and `data/era5_land_hourly_alps_dl_return_levels.rds` from `scripts/5-runoff_marginals.r`.

```r
shiny::runApp("apps/runoff-marginals-explorer")
```

## R Package Demonstration

```{r}
library(rainonsnow)
library(dplyr)
library(probaverse)
```

The R package is titled `rainonsnow` and provides a small distributional learning interface for modelling the distribution of a target variable given some predictors.

For a simple workflow, fit a model with `dl_rqforest()` and then call `predict()`, using the `mtcars` dataset from the stats package.

```{r}
df <- as_tibble(mtcars)
model <- dl_rqforest(
  data = df,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)

df <- mutate(df, distribution = predict(model), .before = everything())
df
```

The predictions are distributions. Take a look at the first distribution using the probaverse, for example, and plot its cdf.

```{r}
plot(df$distribution[[1]], n = 1000)
```

The package also includes a null model to handle failures gracefully. For example, if you ask for `na_action = "null"` and the training data contain missing values, or if the training fails, `dl_rqforest()` returns a `dl_null` object instead of failing:

```{r}
df2 <- as_tibble(mtcars)
df2$wt[2] <- NA_real_

dl_rqforest(
  data = df2,
  yname = "hp",
  xnames = c("wt", "drat", "gear")
)
```

Predicting on these objects always returns a null distribution with the number of rows of the data:

```{r}
predict(dl_null(), newdata = tibble(x = 1:2))
```

## Status

This repository is in an active research/prototyping state, so scripts, paths, and interfaces may change as the workflow evolves.

## License

This repository is licensed under the MIT License - see the `LICENSE` file for details.