---
title: "Rain-on-Snow Distributional Learning Model"
bibliography: analyze.bib
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
---

This document explores a distributional learning model for prediction of gridded runoff from precipitation and snowmelt inputs for a region over the Alps.

## Setup

Load the packages used throughout the analysis and define a simple colour palette for the maps.

```{r}
library(tidyverse)
library(probaverse)
library(terra)
library(tidyterra)
devtools::load_all()

# "Vibrant Summer" palette
# https://coolors.co/palette/ff595e-ffca3a-8ac926-1982c4-6a4c93
pal <- rev(c("#ff595e", "#ffca3a", "#8ac926", "#1982c4", "#6a4c93"))
```

```{r}
dat <- read_rds(here::here("data", "era5_land_hourly_alps_peaks.rds"))
```

Set up two basemap options for the spatial plots:
a static vector basemap from Natural Earth and a tile-based basemap from OpenStreetMap.
The NetCDF stores latitude in `x` and longitude in `y`, so the plotting helpers also convert to the conventional longitude/latitude axes.

```{r}
map_xlim <- range(dat$y, na.rm = TRUE) + c(-0.5, 0.5)
map_ylim <- range(dat$x, na.rm = TRUE) + c(-0.5, 0.5)

map_bbox <- sf::st_bbox(
  c(
    xmin = map_xlim[1],
    xmax = map_xlim[2],
    ymin = map_ylim[1],
    ymax = map_ylim[2]
  ),
  crs = sf::st_crs(4326)
)

world_map <- rnaturalearth::ne_countries(
  scale = 50,
  returnclass = "sf"
) |>
  sf::st_crop(map_bbox)

spatial_basemap <- list(
  geom_sf(
    data = world_map,
    inherit.aes = FALSE,
    fill = NA,
    #color = "grey70",
    linewidth = 1
  ),
  coord_sf(
    xlim = map_xlim,
    ylim = map_ylim,
    expand = FALSE
  ),
  labs(x = "Longitude", y = "Latitude"),
  theme_minimal()
)

spatial_tile_basemap <- list(
  ggspatial::annotation_map_tile(
    type = "osm",
    zoomin = -1,
    progress = "none"
  ),
  coord_sf(
    xlim = map_xlim,
    ylim = map_ylim,
    expand = FALSE
  ),
  labs(x = "Longitude", y = "Latitude"),
  theme_minimal()
)
```


## Exploratory Plots

Cell IDs overtop of the spatial basemap:

```{r}
dat |>
  distinct(cell_id, x, y) |>
  ggplot(aes(y, x)) +
  spatial_basemap +
  geom_tile(fill = "white", colour = "black", alpha = 0.8) +
  geom_text(aes(label = cell_id), size = 2)
```


Spatial runoff at 2 pm on October 5 with a Natural Earth basemap:

```{r}
dat |>
  filter(hour == 14) |>
  ggplot(aes(y, x)) +
  geom_tile(aes(fill = runoff_hourly), alpha = 0.75) +
  spatial_basemap +
  scale_fill_gradientn(colours = pal)
```


Runoff time series by cell (upper-left corner):

```{r}
dat |>
  filter(cell_id == 12) |>
  ggplot(aes(date, runoff_hourly)) +
  geom_point(alpha = 0.75) +
  theme_bw() +
  labs(y = "Hourly Runoff (mm)")
```

POT event timing.

```{r}
dat |>
  filter(cell_id == 12) |>
  ggplot(aes(yday(date), runoff_hourly)) +
  geom_point(alpha = 0.75) +
  theme_bw() +
  labs(y = "Hourly Runoff (mm)")
```

Runoff vs snowmelt by cell (upper-left corner):

```{r}
dat |>
  filter(cell_id == 39) |>
  ggplot(aes(snowmelt_hourly, runoff_hourly)) +
  geom_point(alpha = 0.1) +
  theme_bw()
```

Runoff vs precipitation by cell (upper-left corner).
We're only using precipitation right now for exposition purposes;
rainfall is harder to get and is a task for another day.

```{r}
dat |>
  filter(cell_id == 39) |>
  ggplot(aes(total_precipitation_hourly, runoff_hourly)) +
  geom_point(alpha = 0.1) +
  theme_bw()
```

Probability of snowmelt-rain combination

```{r}
prob_dat <- dat |>
  group_by(cell_id, x, y) |>
  summarise(
    prob_snowyes_rainyes = mean(
      total_precipitation_hourly == 0 & snowmelt_hourly == 0,
      na.rm = TRUE
    ),
    prob_snowyes_rainno = mean(
      total_precipitation_hourly == 0 & snowmelt_hourly > 0,
      na.rm = TRUE
    ),
    prob_snowno_rainyes = mean(
      total_precipitation_hourly > 0 & snowmelt_hourly == 0,
      na.rm = TRUE
    ),
    prob_snowno_rainno = mean(
      total_precipitation_hourly > 0 & snowmelt_hourly > 0,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

Map showing probability of both rainfall and snowmelt occurring together:

```{r}
ggplot(prob_dat, aes(y, x)) +
  spatial_basemap +
  geom_tile(aes(fill = prob_snowyes_rainyes), alpha = 0.85) +
  scale_fill_gradientn(
    "Probability\nRain-on-Snow",
    colours = pal,
    labels = scales::percent_format()
  )
```

2x2 plot showing histograms of probabilities of each combination: snowmelt-rain, no snowmelt-rain, no snowmelt-no rain, snowmelt-no rain:

```{r}
prob_dat |>
  pivot_longer(
    starts_with("prob_"),
    names_to = c(".value", "rain", "snow"),
    names_pattern = "(prob)_snow(.+)_rain(.+)"
  ) |>
  ggplot(aes(prob)) +
  facet_grid(
    rain ~ snow,
    labeller = labeller(
      rain = c("yes" = "Rain", "no" = "No Rain"),
      snow = c("yes" = "Snow", "no" = "No Snow")
    )
  ) +
  geom_histogram() +
  labs(
    caption = paste(
      "Probability that an event has the specified combination of rain and snowmelt\n",
      "pooled across all cells."
    ),
    x = "Probability",
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```



## Distributional Learning Model

Fit a distributional random forest model separately at each grid cell.

This method follows quantile regression forests [@meinshausen2006quantile].
Instead of calculating only the mean at each leaf in the random forest, conditional quantiles are estimated; taken together, these give the full conditional distribution.

Distributional learning wrappers can be found under `R/` in this repository.
These are needed because the base methods (e.g., `quantregForest::quantregForest()`) are geared towards quantiles, not distributions.
The [probaverse](https://probaverse.com) is used for the handling of distributions.

```{r}
models <- dat |>
  nest(data = !c(cell_id, x, y)) |>
  mutate(
    dl_rqforest = map(
      data,
      dl_rqforest,
      yname = "runoff_hourly",
      xnames = c("total_precipitation_hourly", "snowmelt_hourly")
    )
  )

# Inspect
models
```

## Predictions on Training Data

Generate predictive distributions for each hour on record.
At this stage, we are not concerning ourselves with a holdout set, since this work is for expository purposes at this time.

```{r}
pred <- models |>
  mutate(hat = map2(dl_rqforest, data, predict)) |>
  select(!dl_rqforest) |>
  unnest(c(data, hat))

# Inspect
pred
```

Inspect one fitted predictive distribution directly; here's its CDF, for example:

```{r}
example_prediction <- pred$hat[[1]]
plot(example_prediction, n = 1000)
```

Note that the method employed by Bevacqua et al. [@bevacqua2017compound] forces this distribution to be normal.

Here are its quartiles:

```{r}
enframe_quantile(example_prediction, at = c(0.25, 0.5, 0.75))
```

Its mean:

```{r}
mean(example_prediction)
```

Convert the empirical distribution to a generalized Pareto distribution.

```{r}
predgp <- mutate(pred, hat = map(hat, convert_emp_to_gp))
```


## P-P Plot

For distributional learning methods, a P-P plot is a way to assess calibration [@coia2017forecasting].
In other words, could the data have feasibly been drawn from the predictive distributions?
The closer the points are to the 1:1 line, the better calibrated the model is.
Anything "off" from this line indicates a tendency for our predictive distributions to, for example, perhaps be too large, too small, or too wide.
However, this isn't the full picture, because it doesn't reward narrowness of the distribution:
for instance, the marginal distribution would appear perfectly calibrated, but for a predictive model, is too wide.

```{r}
pp_data <- pred |>
  group_by(x, y) |>
  mutate(
    p_model = map2_dbl(hat, runoff_hourly, eval_cdf),
    p_empirical = uscore(p_model)
  ) |>
  select(x, y, starts_with("p_"))

ggplot(pp_data, aes(p_empirical, p_model)) +
  geom_line(aes(group = interaction(x, y)), alpha = 0.1) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    colour = "orange3"
  ) +
  theme_bw()
```

To include both calibration and narrowness of predictions, the quantile score is a useful measure [@gneiting2011making].
Get quantiles and quantile scores from probaverse.

```{r}
qscores_model <- pred |>
  mutate(
    df = map(
      hat,
      enframe_quantile,
      at = 1:99 / 100,
      arg_name = "tau"
    )
  ) |>
  unnest(df) |>
  group_by(x, y, tau) |>
  summarise(
    qscore_model = mean(quantile_score(
      runoff_hourly,
      xhat = quantile,
      tau = tau
    )),
    .groups = "drop"
  )

# Inspect
qscores_model
```

On their own, quantile scores do not hold an obvious interpretation.
They are better when compared against the models obtained from the univariate method -- that is, always predicting the marginal distribution obtained from the univariate data.
This is just like the calculation of $R^2$, except instead of looking at the reduction of variance, it's the reduction in quantile score. Call this the "skill score".

```{r}
null_model <- dat |>
  group_by(x, y) |>
  summarise(runoff_hourly = list(runoff_hourly), .groups = "drop") |>
  mutate(marginal = map(runoff_hourly, dst_empirical))

null_quantiles <- null_model |>
  mutate(
    df = map(marginal, enframe_quantile, at = 1:99 / 100, arg_name = "tau")
  ) |>
  select(!marginal) |>
  unnest(df)

qscores_null <- null_quantiles |>
  mutate(
    qscore_null = pmap_dbl(
      list(runoff_hourly, quantile, tau),
      \(y, q, p) mean(quantile_score(y, xhat = q, tau = p))
    )
  ) |>
  select(x, y, tau, qscore_null)

qscores <- left_join(qscores_null, qscores_model, by = c("x", "y", "tau")) |>
  mutate(skill_score = 1 - qscore_model / qscore_null)

ggplot(qscores, aes(tau, skill_score)) +
  geom_line(
    aes(group = interaction(x, y)),
    alpha = 0.33
  ) +
  labs(
    x = "Quantile Level",
    title = "Skill Score of Quantile Regression Forest Method"
  ) +
  scale_y_continuous("Skill", labels = scales::percent_format()) +
  theme_bw()
```

## Marginal Distributions

Unconditional runoff distributions are useful for calculating extreme events like the 100-year or 200-year runoff event.
The dataset will have to be changed in order to calculate these events, but the process can be demonstrated with these models nonetheless, by [mixing](https://en.wikipedia.org/wiki/Mixture_distribution) all the predictive distributions in each cell.
This calculates the marginal distribution of runoff, unconditional on the model inputs.
Calculations are made easy by the probaverse: just use distplyr's `mix()` function.

```{r}
marginals <- pred |>
  group_by(x, y) |>
  summarise(marginal_runoff = list(mix(hat)), .groups = "drop")
```

Get the marginals from the generalized Pareto distributions.

```{r}
marginalsgp <- predgp |>
  group_by(x, y) |>
  summarise(marginal_runoff = list(mix(hat)), .groups = "drop")
```

Here is an example unconditional distribution (CDF) for one of the cells in the study region, comparing the empirical model and the generalized Pareto model.

```{r}
plot(marginals$marginal_runoff[[4]], n = 1000)
plot(marginalsgp$marginal_runoff[[4]], n = 1000, add = TRUE, col = "red")
legend(
  "bottomright",
  legend = c("Empirical", "Generalized Pareto"),
  col = c("black", "red"),
  lty = 1
)
```

One objective will be to show that, particularly when parametric tails are fit via distributional learning, this estimate of the marginal distribution will surpass that of the univariate method.
I've seen this in small simulations before, and for a binary view of rain and snowmelt processes ([BGC, 2025](https://www.fraserbasin.bc.ca/_Library/TR_Flood/Coldwater%20and%20Nicola%20Rivers%20FFA%20Final%20Report.pdf)), and suspect it's because the shape of the distribution's tail can be explained by the drivers, rather than simply trying to extrapolate the tail from a univariate view. In practical terms: an extreme runoff becomes expected in the face of a large rainfall and large snowmelt, rather than something that appears to have come from "out of the blue".

## Percentiles Maps

Getting the unconditional distributions is useful for making a map of, say, the 100-year flood (with the appropriate dataset).
For now, instead, take a look at the 1% and 99% values.

```{r}
pctl <- marginals |>
  mutate(
    df = map(
      marginal_runoff,
      enframe_quantile,
      at = c(0.01, 0.99),
      arg_name = "tau"
    )
  ) |>
  select(!marginal_runoff) |>
  unnest(df)

ggplot(pctl, aes(y, x)) +
  spatial_basemap +
  facet_wrap(~tau, ncol = 2) +
  geom_tile(aes(fill = quantile), alpha = 0.85) +
  scale_fill_gradientn("Percentile:\nHourly Runoff (mm)", colours = pal)
```

From a more day-to-day prediction standpoint, we can look at the percentiles of the predictive distributions (i.e., the conditional ones), and compare to what was observed.
Do this for 2 pm on October 5; also compare with the mean predicted runoff (i.e., classical machine learning).

```{r}
pctl_hr14 <- pred |>
  filter(hour == 14) |>
  mutate(
    runoff_01 = map_dbl(hat, eval_quantile, at = 0.01),
    runoff_99 = map_dbl(hat, eval_quantile, at = 0.99),
    runoff_mean = map_dbl(hat, mean),
  ) |>
  select(x, y, starts_with("runoff")) |>
  pivot_longer(
    starts_with("runoff"),
    names_to = c(".value", "quantity"),
    names_sep = "_"
  ) |>
  mutate(
    quantity = case_when(
      quantity == "01" ~ "1% Percentile",
      quantity == "99" ~ "99% Percentile",
      quantity == "mean" ~ "Mean",
      TRUE ~ "Observed"
    )
  )

ggplot(pctl_hr14, aes(y, x)) +
  spatial_basemap +
  facet_wrap(~quantity, nrow = 2) +
  geom_tile(aes(fill = runoff), alpha = 0.85) +
  scale_fill_gradientn("Hourly Runoff (mm)", colours = pal)
```

## Return Levels

How many POT events happen per year? This is needed to calculate the return level curve.

```{r}
num_pot_events <- dat |>
  group_by(cell_id, x, y) |>
  summarise(
    num_events_per_year = n() / (diff(range(year(date))) + 1),
    .groups = "drop"
  )
```

First take a look at the return level curve. See that the empirical model plateaus.

```{r}
return_periods <- c(2, 5, 10, 20, 50, 100, 200, 500, 1000)
lvls <- marginals |>
  mutate(
    df = map(
      marginal_runoff,
      enframe_quantile,
      at = return_periods,
      arg_name = "return_period"
    )
  ) |>
  select(!marginal_runoff) |>
  unnest(df)

lvlsgp <- marginalsgp |>
  left_join(num_pot_events, by = c("x", "y")) |>
  mutate(
    df = map(
      marginal_runoff,
      enframe_quantile,
      at = return_periods * num_events_per_year,
      arg_name = "rp_adjusted"
    )
  ) |>
  select(!marginal_runoff) |>
  unnest(df) |>
  mutate(return_period = rp_adjusted / num_events_per_year)

lvls_both <- left_join(
  lvls,
  lvlsgp,
  by = c("x", "y", "return_period"),
  suffix = c("_emp", "_gp")
) |>
  pivot_longer(
    c(return_emp, return_gp),
    names_to = c(".value", "model"),
    names_sep = "_"
  ) |>
  mutate(
    model = case_when(
      model == "emp" ~ "Empirical",
      model == "gp" ~ "Generalized Pareto",
      TRUE ~ model
    )
  )

lvls_both |>
  filter(cell_id == 36) |>
  ggplot(aes(return_period, return, color = model)) +
  geom_line(alpha = 0.75) +
  scale_x_log10("Return Period (years)") +
  scale_color_manual("Model", values = c("black", "red3")) +
  labs(y = "Return Level (mm)") +
  theme_bw()
```



## Rain-snowmelt combinations most likely to produce specified runoff

Get the most likely combination of rain and snowmelt to produce a specified return period runoff event.

First pick the event and get the magnitude.

```{r}
return_period <- 200
grid_mult <- 1.3
q200_events <- marginalsgp |>
  mutate(
    magnitude = map_dbl(
      marginal_runoff,
      eval_return,
      at = .env$return_period,
    )
  ) |>
  select(cell_id, x, y, magnitude)
```

Now get the most likely combination of rain and snowmelt to produce that return level.

```{r}
grids <- dat |>
  nest(data = !c(cell_id, x, y)) |>
  mutate(
    grid = map(
      data,
      grid_from_scatter,
      x = total_precipitation_hourly,
      y = snowmelt_hourly,
      mult = .env$grid_mult
    )
  ) |>
  select(!data) |>
  left_join(q200_events, by = c("x", "y"))

grids_pred <- grids |>
  mutate(
    pred = map2(grid, data, predict),
    pred = map(pred, convert_emp_to_gp)
  )

likely_combinations <- grids_pred |>
  mutate(
    prob_responsible = map2(pred, magnitude, distribution_likeliness)
  ) |>
  unnest(c(pred, prob_responsible)) |>
  group_by(cell_id, x, y)
```

Plot the likely combinations for a given cell:

```{r}
this_cell <- 36
this_data <- filter(dat, cell_id == .env$this_cell)
this_likely <- filter(likely_combinations, cell_id == .env$this_cell)
ggplot(this_likely, aes(total_precipitation_hourly, snowmelt_hourly)) +
  geom_contour(aes(z = prob_responsible))) +
  geom_point(data = this_data, color = "red3") +
  theme_bw()
```